Relational Parametricity for Algebraically Indexed Types

### 2012-06-08 paper/geometry-twocolumn.tex
Latest changes

\section{Motivating Examples}

@@

When writing programs that manipulate geometric data, the basic data
structure is the $n$-tuple of real numbers. In the 2-dimensional case,
tuples $\vec{v} = (x,y)$ may be called upon to represent both
\emph{points}---offsets from some origin---and
\emph{vectors}---offsets in their own right. Points and vectors are
very different despite their common representation.

@@

We now sketch the interpretation of the types $\tyPrim{vec}{e}$ and
$\forall t\mathord:\SynTransl_2.A$. In our calculus, each type has two
interpretations: an index-erasure interpretation that ignores the
indexing expression $e$, and a relational interpretation as a binary
relation on the index-erasure interpretation. For now, we assume we
are given some way of turning an expression $e$ of sort $\SynTransl_2$
into an actual element $\sem{e}\rho$ of the group $\Transl_2$, under
some environment $\rho$ which tells how to interpret the free
variables. The (simplified) index-erasure and relation interpretations
of the types $\tyPrim{vec}{e}$ are:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\tyPrim{vec}{e}} & \mathbb{R}^2 \\
    \rsem{\tyPrim{vec}{e}}{}\rho & R_{\sem{t}\rho} = \{ (\vec{v},\vec{v'}) \sepbar \vec{v'} = \vec{v} + \sem{t}\rho \}
  \end{array}
\end{displaymath}
The index-erasure interpetation of the universal quantification
$\forall t\mathord:\SynTransl_2.A$ is simply the interpretation of the
underlying type $A$. The relational interpretation intersects the
relational interpretations of the type $A$ under all extensions of the
relational environment $\rho$:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\forall t\mathord:\SynTransl_2.A} & \tySem{A} \\
    \rsem{\forall t\mathord:\SynTransl_2.A}{}\rho & \bigcap\{ \rsem{A}{\relEnv{E}}\rho' \sepbar \rho' \in \extends{\rho}{i\mathord:\SynTransl_2} \}
  \end{array}
\end{displaymath}
The exact definition of $\extends{\rho}{i\mathord:\SynTransl_2}$
depends on which (potentially non-compositional) interpretation of
relational environments that we take. For now, it will suffice to
think of relational environments as simply mappings from free
variables to the group $\Transl_2$, and extension as just adding an
extra component to the tuple. The general index-erasure semantics will
be given in \autoref{sec:index-erasure-semantics}. The full relational
interpretation is in general more complex, due to the possibility of
non-compositional interpretations of free index variables. The
relational interpretation will be presented in
\autoref{sec:relational-semantics}.

@@

Invariance under change of origin is the key feature of affine
geometry, and the central operation of affine geometry is the affine
combination of points: $\lambda_1\vec{v_1} + \lambda_2\vec{v_2}$,
where $\lambda_1 + \lambda_2 = 1$. Geometrically, affine combination
can be interpreted as interpolation between the points represented by
$\vec{v_1}$ and $\vec{v_2}$. We add affine combination of points to
our calculus, with the following typing and intended denotation:
\begin{displaymath}
  \begin{array}{l}
    \mathrm{affComb} :\forall t \mathord: \SynTransl_2.\ \tyPrim{vec}{t} \to \tyPrimNm{real} \to \tyPrim{vec}{t} \to \tyPrim{vec}{t} \\
    \sem{\mathrm{affComb}}\ \vec{v_1}\ r\ \vec{v_2} = (1-t)\vec{v_1} + t\vec{v_2}
  \end{array}
\end{displaymath}
It can be verified by hand that the intended denotation
$\sem{\mathrm{affComb}}$ has the invariance property specified by its
type.  By defining the primitive function $\mathrm{affComb}$ to take a
single $\tyPrimNm{real}$ parameter $t$, we can easily ensure that we
are taking the affine combination of two representatives of points,
and not just an arbitrary linear combination.

\begin{example}
  The evaluation of quadratic B\'{e}zier curves (B\'{e}zier curves
  with two endpoints and a single control point) can be expressed
  using the affine combination primitive and three steps of De
  Casteljau's algorithm:
  \begin{displaymath}
    \begin{array}{@{}l}
      \mathrm{quadB\acute{e}zier} : \forall t \mathord:\SynTransl_2.\ \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrimNm{real} \to \tyPrim{vec}{t} \\
      \mathrm{quadB\acute{e}zier}\ [t]\ p_0\ p_1\ p_2\ s = \\
      \quad \mathrm{affComb}\ [F]\ (\mathrm{affComb}\ [F]\ p_0\ s\ p_1)\ s\ (\mathrm{affComb}\ [F]\ p_1\ s\ p_2) \\
    \end{array}
  \end{displaymath}
  For two endpoints $p_0$ and $p_2$ and a control point $p_1$, an
  application $\mathrm{quadB\acute{e}zier}\ p_0\ p_1\ p_2\ s$ for $s
  \in [0,1]$ gives us the point on the curve at ``time'' $s$. Note
  that the type of $\mathrm{quadB\acute{e}zier}$ tells us immediately
  that it preserves all changes of origin.
\end{example}

Affine combination only provides us with an operation on points. We
are also interested in operations on the vectors representing offsets
between points. We now examine the correct types to assign to the
vector space operations of addition of vectors, negation of vectors,
multiplication by a scalar and the zero vector. The typings of these
operations will make use of the abelian group structure of change of
origin translations.

@@

From basic linear algebra, we know that the vector operations of
addition, negation and scaling, while not invariant under change of
origin, are invariant under change of basis. Following the same
methodology as we did for origin invariance, we can express basis
invariance by means of preservation of relations. For origin
invariance, we quantified over members of the group $\SynTransl_2$ of
$2$-dimensional translations. Members of this group represent changes
of origin. Change of basis is accomplished by applying an invertible
linear map, and the collection of all invertible linear maps on
$\mathbb{R}^2$ forms the group $\GL_2$.

@@

@@

FIXME: we now have all the operations of affine geometry, along with
the property that they are invariant under affine transformations.

\paragraph{Euclidean Geometry} Euclidean geometry extends affine
geometry with the operation of \emph{dot product} (also known as the
\emph{inner product}) of two vectors, which produces a real
number. The standard Euclidean dot product is defined on the
coordinate representation in the following way:
\begin{displaymath}
  (x_1,y_1) \cdot (x_2,y_2) = x_1x_2 + y_1y_2
\end{displaymath}
The dot product is used to define properties of vectors such as their
\emph{norm} $||\vec{v}|| = \vec{v}\cdot\vec{v}$, which is the square
of the length of the vector $\vec{v}$, and the notion of
orthogonality: two vectors $\vec{v_1}$ and $\vec{v_2}$ are orthogonal
if $\vec{v_1}\cdot\vec{v_2} = 0$.

To incorporate the operation of dot product into our calculus, we must
give it a type. It is not the case that the dot product is invariant
under the action of either of the groups $\GL_2$ or $\Transl_2$ that
we have considered so far. In fact, the dot product is invariant under
the subgroup $\Orth_2$ of $\GL_2$ of \emph{orthogonal} linear
transformations. The group $\Orth_2$ consists of invertible linear
maps whose determinant is either $1$ or $-1$. We introduce orthogonal
transformations as a new sort $\Orth_2$, and overload the
multiplicative group operations syntax for inhabitants of
$\Orth_2$. We also assume a function $\iota_O$ that takes $e :
\Orth_2$ to $\iota_O(e) : \SynGL_2$. We may now assign the following type
to the dot product:
\begin{displaymath}
  (\cdot) : \forall O \mathord: \Orth(2).\ \tyPrim{vec}{\iota_O(O), 0} \to \tyPrim{vec}{\iota_O(O), 0} \to \tyPrimNm{real}
\end{displaymath}

\begin{figure}[t]
  \centering
  \begin{eqnarray*}
    0   &:& \forall s \mathord:\Scale.\ \tyPrim{real}{s} \\
    1   &:& \tyPrim{real}{1} \\
    (+) &:& \forall s \mathord:\Scale.\ \tyPrim{real}{s} \to \tyPrim{real}{s} \to \tyPrim{real}{s} \\
    (-) &:& \forall s \mathord:\Scale.\ \tyPrim{real}{s} \to \tyPrim{real}{s} \to \tyPrim{real}{s} \\
    (*) &:& \forall s_1,s_2 \mathord:\Scale.\ \tyPrim{real}{s_1} \to \tyPrim{real}{s_2} \to \tyPrim{real}{s_1s_2} \\
    (/) &:&
    \begin{array}[t]{@{}l}
      \forall s_1,s_2 \mathord:\Scale.\ \tyPrim{real}{s_1} \to \tyPrim{real}{s_2} \\
      \hspace{3.5cm}\to \tyPrim{real}{s_1s_2^{-1}} + \tyPrimNm{unit} \\
    \end{array}\\
    \mathrm{abs} &:& \forall s \mathord:\Scale.\ \tyPrim{real}{s} \to \tyPrim{real}{|s|} %\\
%    \mathrm{sqrt} &:& \tyPrim{real}{1} \to \tyPrim{real}{1}
  \end{eqnarray*}
  \caption{Operations on scaled real numbers}
  \label{fig:real-ops}
\end{figure}

The cross product is another operation that takes pairs of vectors and
produces a real number. Algebraically, the cross product is defined on
coordinate representations as follows:
\begin{displaymath}
  (x_1,y_1) \times (x_2,y_2) = x_1y_2 - x_2y_1
\end{displaymath}
Geometrically, the cross product of two vectors is the signed area of
the parallelogram described by the pair of input vectors. It turns out
that the value of the cross product of two vectors varies with the
determinant of the matrix describing the basis of the vector space. We
therefore augment our calculus with a new sort $\Scale$ of scale
factors. Semantically, $\Scale$ ranges over the non-zero real numbers
and forms an abelian group, which we write multiplicatively. We also
add two new operations: determinant, $\det B$, which takes an
inhabitant of $\SynGL_2$ to its determinant in $\Scale$, and the
operation of absolute value that takes scaling factors to scaling
factors which we write as $|e|$. We also refine the type
$\tyPrimNm{real}$ of real numbers to take a parameter of sort
$\Scale$: $\tyPrim{real}{e}$, where the old type $\tyPrimNm{real}$ is
just $\tyPrim{real}{1}$. The collection of operations on real numbers
with scaling factors is shown in \autoref{fig:real-ops}.

With the additional sort $\Scale$, and the refined type of real
numbers, we can state the type of the cross product.
\begin{displaymath}
  (\times) : \forall B \mathord: \SynGL_2.\ \tyPrim{vec}{B,0} \to \tyPrim{vec}{B,0} \to \tyPrim{real}{\det B}
\end{displaymath}
For orthogonal transformations $O : \Orth_2$, the absolute value of
the determinant is always $1$, so we assume the equation $|\det O| =
1$ to hold.

\begin{example}
  Using the structure we have defined in this subsection, we can
  define the following function that computes the area of a
  triangle. By the type of this function, we can see that the area of
  a triangle is invariant under orthognal transformations and
  arbitrary translations.
  \begin{displaymath}
    \begin{array}{@{}l}
      \mathrm{area} : \forall O\mathord:\Orth_2, t\mathord:\SynTransl_2.\ \\
      \hspace{0.8cm}\tyPrim{vec}{\iota_OO, t} \to \tyPrim{vec}{\iota_OO, t} \to \tyPrim{vec}{\iota_OO, t} \to \tyPrim{real}{1} \\
      \mathrm{area}\ [O]\ [t]\ p_1\ p_2\ p_3 = \frac{1}{2} * \mathrm{abs}\ ((p_2 - p_1) \times (p_3 - p_1))
    \end{array}
  \end{displaymath}
  In the body of this function, we have performed the calculation in
  several steps, each of which removes some of the symmetry described
  by the type of $\mathrm{area}$. First we have computed the two offset
  vectors $p_2 - p_1$ and $p_3 - p_1$. These operations remove the
  effect of translations on the result, in exactly the same way as the
  type isomorphism we demonstrated above. We then compute the cross
  product of the two vectors to determine the area of the
  parallelogram described by the sides of the triangle. This value has
  type $\mathsf{real}\langle \det O \rangle$. We know that the
  determinant of an orthogonal transformation is always either $-1$ or
  $1$, so we have a result which varies according to possible
  reflections of the space. To remove this variation, we apply the
  $\mathrm{abs}$ function to leave us with a value of type
  $\mathsf{real}\langle |\det O| \rangle$. This type is equal to the
  type $\mathsf{real}\langle 1 \rangle$. Finally, we multiply by
  $\frac{1}{2}$ to recover the area of the triangle, rather than the
  whole parallelogram.
\end{example}

@@

@@

>>>>

\section{Motivating Examples}
\label{sec:motivating-examples}

@@

When writing programs that manipulate geometric data, the basic data
structure is the $n$-tuple of numbers. In the 2-dimensional case,
tuples $\vec{v} = (x,y)$ are called upon to represent both
\emph{points}---offsets from some origin---and
\emph{vectors}---offsets in their own right. Points and vectors are
very different despite their common representation.

@@

We now sketch the interpretation of the types $\tyPrim{vec}{e}$ and
$\forall t\mathord:\SynTransl_2.A$. In our calculus, each type has two
interpretations: an index-erasure interpretation that ignores the
indexing expression $e$, and a relational interpretation as a binary
relation on the index-erasure interpretation. For now, we assume we
are given some way of turning an expression $e$ of sort $\SynTransl_2$
into an actual element $\sem{e}\rho$ of the group $\Transl_2$, under
some environment $\rho$ which tells how to interpret the free
variables. The (simplified) index-erasure and relation interpretations
of the types $\tyPrim{vec}{e}$ are:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\tyPrim{vec}{e}} & \mathbb{R}^2 \\
    \rsem{\tyPrim{vec}{e}}{}\rho & R_{\sem{t}\rho} = \{ (\vec{v},\vec{v'}) \sepbar \vec{v'} = \vec{v} + \sem{e}\rho \}
  \end{array}
\end{displaymath}
The index-erasure interpetation of universal quantification $\forall
t\mathord:\SynTransl_2.A$ is simply the interpretation of the
underlying type $A$. The relational interpretation intersects the
relational interpretations of the type $A$ under all extensions of the
relational environment $\rho$:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\forall t\mathord:\SynTransl_2.A} & \tySem{A} \\
    \rsem{\forall t\mathord:\SynTransl_2.A}{}\rho & \bigcap\{ \rsem{A}{}{\rho'} \sepbar \rho' \in \extends{\rho}{i\mathord:\SynTransl_2} \}
  \end{array}
\end{displaymath}
The exact definition of $\extends{\rho}{i\mathord:\SynTransl_2}$
depends on which (possibly non-compositional) interpretation of
relational environments that we take. For now, it will suffice to
think of relational environments as simply mappings from free
variables to the group $\Transl_2$, and extension as just adding an
extra component to the tuple. The general index-erasure semantics will
be given in \autoref{sec:index-erasure-semantics}. The full relational
interpretation is in general more complex, due to the possibility of
non-compositional interpretations of free index variables. The
relational interpretation will be presented in
\autoref{sec:relational-semantics}.

@@

Invariance under change of origin is the key feature of affine
geometry, and the central operation of affine geometry is the affine
combination of points: $\lambda_1\vec{v_1} + \lambda_2\vec{v_2}$,
where $\lambda_1 + \lambda_2 = 1$. Geometrically, affine combination
can be interpreted as interpolation between the points represented by
$\vec{v_1}$ and $\vec{v_2}$. We add affine combination of points to
our calculus, with the following typing and intended denotation:
\begin{displaymath}
  \begin{array}{l}
    \mathrm{affComb} :\forall t \mathord: \SynTransl_2.\ \tyPrim{vec}{t} \to \tyPrimNm{real} \to \tyPrim{vec}{t} \to \tyPrim{vec}{t} \\
    \sem{\mathrm{affComb}}\ \vec{v_1}\ r\ \vec{v_2} = (1-r)\vec{v_1} + r\vec{v_2}
  \end{array}
\end{displaymath}
It can be verified by hand that the intended denotation
$\sem{\mathrm{affComb}}$ has the invariance property specified by its
type.
% By defining the primitive function $\mathrm{affComb}$ to take a
% single $\tyPrimNm{real}$ parameter $t$, we can easily ensure that we
% are taking the affine combination of two representatives of points,
% and not just an arbitrary linear combination.

\begin{example}
  The evaluation of quadratic B\'{e}zier curves (B\'{e}zier curves
  with two endpoints and a single control point) can be expressed
  using the affine combination primitive and three steps of De
  Casteljau's algorithm:
  \begin{displaymath}
    \begin{array}{@{}l}
      \mathrm{quadB\acute{e}zier} : \forall t \mathord:\SynTransl_2.\ \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrimNm{real} \to \tyPrim{vec}{t} \\
      \mathrm{quadB\acute{e}zier}\ [t]\ p_0\ p_1\ p_2\ s = \\
      \quad \mathrm{affComb}\ [t]\ (\mathrm{affComb}\ [t]\ p_0\ s\ p_1)\ s\ (\mathrm{affComb}\ [t]\ p_1\ s\ p_2) \\
    \end{array}
  \end{displaymath}
  For two endpoints $p_0$ and $p_2$ and a control point $p_1$, an
  application $\mathrm{quadB\acute{e}zier}\ p_0\ p_1\ p_2\ s$ for $s
  \in [0,1]$ gives us the point on the curve at ``time'' $s$. Note
  that the type of $\mathrm{quadB\acute{e}zier}$ tells us immediately
  that it preserves all changes of origin.
\end{example}

% Affine combination only provides us with an operation on points. We
% are also interested in operations on the vectors representing offsets
% between points. We now examine the correct types to assign to the
% vector space operations of addition of vectors, negation of vectors,
% multiplication by a scalar and the zero vector. The typings of these
% operations will make use of the abelian group structure of change of
% origin translations.

\fixme{Need to make clear that the point is to make use of the
  algebraic indexing structure}

@@

From basic linear algebra, we know that the vector operations of
addition, negation and scaling, while not invariant under change of
origin, are invariant under change of basis. Following the same
methodology as we did for origin invariance, we can express basis
invariance by means of preservation of relations indexed by changes of
basis. Change of basis is accomplished by applying an invertible
linear map, and the collection of all invertible linear maps on
$\mathbb{R}^2$ forms the group $\GL_2$.

@@

\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\tyPrim{vec}{e_B,e_t}} & \mathbb{R}^2 \\
    \rsem{\tyPrim{vec}{e_B,e_t}}{}\rho & \{ (\vec{v},\vec{v'}) \sepbar \vec{v'} = \sem{e_B}\rho\vec{v} + \sem{e_t}\rho \}
  \end{array}
\end{displaymath}

@@

\fixme{we now have all the operations of affine geometry, along with
  the property that they are invariant under affine transformations.}

\paragraph{Euclidean Geometry} Euclidean geometry extends affine
geometry with the operation of \emph{dot product} (also known as the
\emph{inner product}) of two vectors, which produces a real
number. The standard Euclidean dot product is defined on the
coordinate representation in the following way:
\begin{displaymath}
  (x_1,y_1) \cdot (x_2,y_2) = x_1x_2 + y_1y_2
\end{displaymath}
% The dot product is used to define properties of vectors such as their
% \emph{norm} $||\vec{v}|| = \vec{v}\cdot\vec{v}$, which is the square
% of the length of the vector $\vec{v}$, and the notion of
% orthogonality: two vectors $\vec{v_1}$ and $\vec{v_2}$ are orthogonal
% if $\vec{v_1}\cdot\vec{v_2} = 0$.

To incorporate the operation of dot product into our calculus, we must
give it a type. It is not the case that the dot product is invariant
under the action of either of the groups $\GL_2$ or $\Transl_2$ that
we have considered so far. In fact, the dot product is invariant under
the subgroup $\Orth_2$ of $\GL_2$ of \emph{orthogonal} linear
transformations. The group $\Orth_2$ consists of invertible linear
maps whose determinant is either $1$ or $-1$. We introduce orthogonal
transformations as a new sort $\SynOrth_2$, and overload the
multiplicative group operations syntax for inhabitants of
$\SynOrth_2$. We also assume a function $\iota_O$ that takes $e :
\SynOrth_2$ to $\iota_O(e) : \SynGL_2$. We may now assign the following type
to the dot product:
\begin{displaymath}
  (\cdot) : \forall O \mathord: \SynOrth(2).\ \tyPrim{vec}{\iota_OO, 0} \to \tyPrim{vec}{\iota_OO, 0} \to \tyPrimNm{real}
\end{displaymath}

\begin{figure}[t]
  \centering
  \begin{eqnarray*}
    0   &:& \forall s \mathord:\SynGL_1.\ \tyPrim{real}{s} \\
    1   &:& \tyPrim{real}{1} \\
    (+) &:& \forall s \mathord:\SynGL_1.\ \tyPrim{real}{s} \to \tyPrim{real}{s} \to \tyPrim{real}{s} \\
    (-) &:& \forall s \mathord:\SynGL_1.\ \tyPrim{real}{s} \to \tyPrim{real}{s} \to \tyPrim{real}{s} \\
    (*) &:& \forall s_1,s_2 \mathord:\SynGL_1.\ \tyPrim{real}{s_1} \to \tyPrim{real}{s_2} \to \tyPrim{real}{s_1s_2} \\
    (/) &:&
    \begin{array}[t]{@{}l@{}l}
      \forall s_1,s_2 \mathord:\SynGL_1.\ \tyPrim{real}{s_1}\ & \to \tyPrim{real}{s_2} \\
      &\to \tyPrim{real}{s_1s_2^{-1}} + \tyPrimNm{unit} \\
    \end{array}\\
    \mathrm{abs} &:& \forall s \mathord:\SynGL_1.\ \tyPrim{real}{s} \to \tyPrim{real}{|s|} %\\
%    \mathrm{sqrt} &:& \tyPrim{real}{1} \to \tyPrim{real}{1}
  \end{eqnarray*}
  \caption{Operations on scaled real numbers}
  \label{fig:real-ops}
\end{figure}

The cross product of two vectors is defined on coordinate
representations as follows:
\begin{displaymath}
  (x_1,y_1) \times (x_2,y_2) = x_1y_2 - x_2y_1
\end{displaymath}
Geometrically, the cross product is the signed area of the
parallelogram described by the pair of input vectors. It turns out
that the value of the cross product of two vectors varies with the
determinant of the matrix describing the basis of the vector space. We
therefore augment our calculus with a new sort $\SynGL_1$ of scale
factors. Semantically, $\SynGL_1$ ranges over the non-zero real
numbers and forms an abelian group, which we write
multiplicatively. We also add two new operations: determinant, $\det
B$, which takes an inhabitant of $\SynGL_2$ to its determinant in
$\SynGL_1$, and the operation of absolute value that takes scaling
factors to scaling factors which we write as $|e|$. We also refine the
type $\tyPrimNm{real}$ of real numbers to take a parameter of sort
$\SynGL_1$: $\tyPrim{real}{e}$, where the old type $\tyPrimNm{real}$
is just $\tyPrim{real}{1}$. The collection of operations on real
numbers with scaling factors is shown in \autoref{fig:real-ops}.

With the additional sort $\SynGL_1$, and the refined type of real
numbers, we can state the type of the cross product.
\begin{displaymath}
  (\times) : \forall B \mathord: \SynGL_2.\ \tyPrim{vec}{B,0} \to \tyPrim{vec}{B,0} \to \tyPrim{real}{\det B}
\end{displaymath}
For orthogonal transformations $O : \Orth_2$, the absolute value of
the determinant is always $1$, so we assume the equation $|\det
(\iota_O O)| = 1$ to hold.

\begin{example}
  Using the structure we have defined in this subsection, we can
  define the following function that computes the area of a
  triangle. By the type of this function, we can see that the area of
  a triangle is invariant under orthogonal transformations and
  arbitrary translations \fixme{i.e. isometries}.
  \begin{displaymath}
    \begin{array}{@{}l}
      \mathrm{area} : \forall O\mathord:\SynOrth_2, t\mathord:\SynTransl_2.\ \\
      \hspace{0.8cm}\tyPrim{vec}{\iota_OO, t} \to \tyPrim{vec}{\iota_OO, t} \to \tyPrim{vec}{\iota_OO, t} \to \tyPrim{real}{1} \\
      \mathrm{area}\ [O]\ [t]\ p_1\ p_2\ p_3 = \frac{1}{2} * \mathrm{abs}\ ((p_2 - p_1) \times (p_3 - p_1))
    \end{array}
  \end{displaymath}
  In the body of this function, we have performed the calculation in
  several steps, each of which removes some of the symmetry described
  by the type of $\mathrm{area}$. First we have computed the two
  offset vectors $p_2 - p_1$ and $p_3 - p_1$. These operations remove
  the effect of translations on the result, in exactly the same way as
  the type isomorphism we demonstrated above \fixme{ref}. We then
  compute the cross product of the two vectors to determine the area
  of the parallelogram described by the sides of the triangle. This
  value has type $\tyPrim{real}{\det (\iota_O O)}$. We know that the
  determinant of an orthogonal transformation is always either $-1$ or
  $1$, so we have a result which varies according to possible
  reflections of the space. To remove this variation, we apply the
  $\mathrm{abs}$ function to leave us with a value of type
  $\tyPrim{real}{|\det (\iota_O O)|}$. This type is equal to the type
  $\tyPrim{real}{1}$. Finally, we multiply by $\frac{1}{2}$ to recover
  the area of the triangle, rather than the whole parallelogram.
\end{example}

@@

Need a function $\iota_1 : \SynGL_1 \to \SynGL_2$. Such that $\det (\iota_1 s) = s^2$

@@

Blah blah blah. Need to work this out in terms of representation indepdence.

### 2012-06-08 paper/paper.tex
Half way through updating the geometry example
Add \fixme newcommand
Substantial changes to the motivating examples section
Latest changes

\usepackage[usenames]{color}
\definecolor{citationcolour}{rgb}{0,0.4,0.2}
\definecolor{linkcolour}{rgb}{0,0,0.8}
\usepackage{hyperref}
\hypersetup{colorlinks=true,
            urlcolor=linkcolour,
            linkcolor=linkcolour,
            citecolor=citationcolour,
            pdftitle=Relational Parametricity for Algebraically-Indexed Types,
            pdfauthor={Robert Atkey, Neil Ghani, Patricia Johann, Andrew Kennedy},
            pdfkeywords={}}  
\def\sectionautorefname{Section}
\def\subsectionautorefname{Section}

@@

\newcommand{\GA}{\mathrm{GA}}
\newcommand{\GL}{\mathrm{GL}}
\newcommand{\Is}{\mathrm{Is}}
\newcommand{\Orth}{\mathrm{O}}
\newcommand{\SOrth}{\mathrm{SO}}
\newcommand{\Transl}{\mathrm{T}}
\newcommand{\Scale}{\mathrm{Scale}}

@@

\newcommand{\tyArr}{\to}
\newcommand{\tyProduct}{\times}
\newcommand{\tyX}[1]{\texttt{X}\langle #1 \rangle}
\newcommand{\isType}{\textsf{ type}}
\newcommand{\isCtxt}{\textsf{ ctxt}}
\newcommand{\Ty}{\textsf{Ty}}
\newcommand{\ty}{\textsf{ty}}
\newcommand{\subst}[2]{\mathrm{Subst}(#1,#2)}

@@

\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{example}{Example}
\newcommand{\lemref}[1]{\hyperref[#1]{Lemma \ref*{#1}}}
\newcommand{\thmref}[1]{\hyperref[#1]{Theorem \ref*{#1}}}
\newcommand{\propref}[1]{\hyperref[#1]{Proposition \ref*{#1}}}
\newcommand{\corref}[1]{\hyperref[#1]{Corollary \ref*{#1}}}
\newcommand{\conref}[1]{\hyperref[#1]{Conjecture \ref*{#1}}}
\newcommand{\exref}[1]{\hyperref[#1]{Example \ref*{#1}}}
\newcommand{\statementref}[1]{\hyperref[#1]{Statement \ref*{#1}}}

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\input{section-introduction.tex}
\input{section-motivating-examples.tex}
\input{section-background.tex}
\input{section-general-framework.tex}
\input{section-instantiations.tex}
\input{section-discussion.tex}

>>>>

\usepackage[usenames]{color}
\definecolor{citationcolour}{rgb}{0,0.4,0.2}
\definecolor{linkcolour}{rgb}{0,0,0.8}
\usepackage{hyperref}
\hypersetup{colorlinks=true,
            urlcolor=linkcolour,
            linkcolor=linkcolour,
            citecolor=citationcolour,
            pdftitle=Relational Parametricity for Algebraically-Indexed Types,
            pdfauthor={Robert Atkey, Neil Ghani, Patricia Johann, Andrew Kennedy},
            pdfkeywords={}}  
\def\sectionautorefname{Section}
\def\subsectionautorefname{Section}
\def\subsubsectionautorefname{Section}

@@

\newcommand{\GA}{\mathrm{GA}}
\newcommand{\GL}{\mathrm{GL}}
\newcommand{\Is}{\mathrm{Is}}
\newcommand{\Orth}{\mathrm{O}}
\newcommand{\SOrth}{\mathrm{SO}}
\newcommand{\Transl}{\mathrm{T}}
\newcommand{\SynTransl}{\mathsf{T}}
\newcommand{\SynGL}{\mathsf{GL}}
\newcommand{\SynOrth}{\mathsf{O}}
\newcommand{\Scale}{\mathrm{Scale}}

@@

\newcommand{\tyArr}{\to}
\newcommand{\tyProduct}{\times}
\newcommand{\tyX}[1]{\texttt{X}\langle #1 \rangle}
\newcommand{\isType}{\textsf{ type}}
\newcommand{\isCtxt}{\textsf{ ctxt}}
\newcommand{\Ty}{\textsf{Ty}}
\newcommand{\ty}{\textsf{ty}}

@@

\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newcommand{\lemref}[1]{\hyperref[#1]{Lemma \ref*{#1}}}
\newcommand{\thmref}[1]{\hyperref[#1]{Theorem \ref*{#1}}}
\newcommand{\propref}[1]{\hyperref[#1]{Proposition \ref*{#1}}}
\newcommand{\corref}[1]{\hyperref[#1]{Corollary \ref*{#1}}}
\newcommand{\conref}[1]{\hyperref[#1]{Conjecture \ref*{#1}}}
\newcommand{\exref}[1]{\hyperref[#1]{Example \ref*{#1}}}
\newcommand{\statementref}[1]{\hyperref[#1]{Statement \ref*{#1}}}

@@

\newtheoremstyle{examplestyle}
  {\topsep}  % space above
  {\topsep}  % space below
  {\normalfont}% name of font to use in the body of the theorem
  {0em}% measure of space to indent
  {\bfseries}% name of head font
  {.}% punctuation between head and body
  {5pt plus 1pt minus 1pt}% space after theorem head; " " = normal interword space
  {}% Manually specify head
\theoremstyle{examplestyle}
\newtheorem{example}{Example}

\newtheoremstyle{restatementstyle}
  {\topsep}  % space above
  {\topsep}  % space below
  {\itshape}% name of font to use in the body of the theorem
  {0em}% measure of space to indent
  {\bfseries}% name of head font
  {.}% punctuation between head and body
  {5pt plus 1pt minus 1pt}% space after theorem head; " " = normal interword space
  {\thmname{#1} \thmnote{(Restatement of #3)}}% Manually specify head
\theoremstyle{restatementstyle}
\newtheorem{restateLemma}{Lemma}
\newtheorem{restateTheorem}{Theorem}

\newcommand{\fixme}[1]{\textbf{FIXME: #1}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\input{section-introduction.tex}
%\input{section-motivating-examples.tex}
\input{geometry-twocolumn.tex}
\input{section-background.tex}
\input{section-general-framework.tex}
\input{section-instantiations.tex}
\input{section-discussion.tex}
